Road Segment Prioritization for Bicycle Infrastructure

### 2020-10-30 README.Rmd
paper updates and formatting from latex -> rmd

### Hello

@@

Segregated cycling infrastructure\footnote{Segregated cycling infrastructure refers to road space that is allocated to cyclists only, with physical separation to protect cyclists from other modes of transport.} has been shown to increase cycling uptake [@aldred2019impacts; @goodman2014new; @marques2015infrastructure], with the separation from motorized vehicles being key [@winters2011motivators]. Revealed preference of cyclists shows that they are willing to deviate from the most efficient routes in order to commute on safer roads [@crane2017longitudinal]. However, such deviations are only considered if they do not considerably increase route circuitry; behaviour studies have found that the probability of choosing a route decreases in proportion to its length relative to the shortest route [@broach2011bicycle; @winters2010far]. Another defining feature for cycling infrastructure is how well connected it is. Cyclists prefer cohesive infrastructure, particularly when cycling on arterial roads with high levels of motorized traffic [@stinson2003commuter], and the lack of well-connected cycling infrastructure is one of the main obstacles to increasing cycling uptake [@caulfield2012determining]. While direct and cohesive cycling networks have been shown to positively impact cycling rates, density\footnote{making an area's bicycle network denser means adding more cycling routes in the area and thereby giving cyclists more route options} of the cycling network is also vital [@schoner2014missing].

@@

These problem formulations do not explicitly solve for continuity, which is dealt with using a either (a) a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], (b) a constraint on deviation from shortest paths [@duthie2014optimization], or (c) a discontinuity penalty [@mauttone2017bicycle]. To solve for continuity, the graph-theoretic concept of *connected components*, has been used. @natera2019data study the existing cycling network in terms of its disconnected components and introduce two different algorithms to connect these components by their most critical links\footnote{\textit{link} refers to a road segment throughout this research} and, in doing so, measure the size of the growth of the largest connected component as a function of the kilometers of network added. They observe that small investments at strategic points have a large impact on connectivity in most cases. The concept of connected components is also at the core of the methodology proposed by @olmos2020data. After routing the cycling demand onto the network links, they use percolation theory to filter out the links based on the aggregate flow\footnote{\textit{flow} is used throughout this research to refer to the cycling demand when it is routed onto the road network. The flow on any road segment is the cumulative demand on it, resulting from cyclists commuting between various OD pairs} passing through them. They vary the flow threshold for filtering to identify the minimum flow at which the whole city is connected by a giant component. The results show a cycling network that connects the entire city, and subtracting links intersecting with current cycling infrastructure identifies links proposed for intervention.

@@

@simini2012

@@

```{r potdemhistograms, fig.cap="Current and Potential Cycling Demand", out.width="80%", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/desire_facet_cycling.png")
```

@@

@@

The weighting profiles are therefore adjusted to favor less-stressful streets (based on information from Table \ref{table:osmroadtypes}), and roads with existing cycling infrastructure. This is also in line with the creation of LTNs, as residential streets are those where motorized traffic is most likely to be banned in the creation of LTNs.

>>>>

@@

Segregated cycling infrastructure^[Segregated cycling infrastructure refers to road space that is allocated to cyclists only, with physical separation to protect cyclists from other modes of transport.] has been shown to increase cycling uptake [@aldred2019impacts; @goodman2014new; @marques2015infrastructure], with the separation from motorized vehicles being key [@winters2011motivators]. Revealed preference of cyclists shows that they are willing to deviate from the most efficient routes in order to commute on safer roads [@crane2017longitudinal]. However, such deviations are only considered if they do not considerably increase route circuitry; behaviour studies have found that the probability of choosing a route decreases in proportion to its length relative to the shortest route [@broach2011bicycle; @winters2010far]. Another defining feature for cycling infrastructure is how well connected it is. Cyclists prefer cohesive infrastructure, particularly when cycling on arterial roads with high levels of motorized traffic [@stinson2003commuter], and the lack of well-connected cycling infrastructure is one of the main obstacles to increasing cycling uptake [@caulfield2012determining]. While direct and cohesive cycling networks have been shown to positively impact cycling rates, density^[making an area's bicycle network denser means adding more cycling routes in the area and thereby giving cyclists more route options} of the cycling network is also vital [@schoner2014missing]].

@@

These problem formulations do not explicitly solve for continuity, which is dealt with using a either (a) a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], (b) a constraint on deviation from shortest paths [@duthie2014optimization], or (c) a discontinuity penalty [@mauttone2017bicycle]. To solve for continuity, the graph-theoretic concept of *connected components*, has been used. @natera2019data study the existing cycling network in terms of its disconnected components and introduce two different algorithms to connect these components by their most critical links^[*link* refers to a road segment throughout this research] and, in doing so, measure the size of the growth of the largest connected component as a function of the kilometers of network added. They observe that small investments at strategic points have a large impact on connectivity in most cases. The concept of connected components is also at the core of the methodology proposed by @olmos2020data. After routing the cycling demand onto the network links, they use percolation theory to filter out the links based on the aggregate flow^[*flow* is used throughout this research to refer to the cycling demand when it is routed onto the road network. The flow on any road segment is the cumulative demand on it, resulting from cyclists commuting between various OD pairs] passing through them. They vary the flow threshold for filtering to identify the minimum flow at which the whole city is connected by a giant component. The results show a cycling network that connects the entire city, and subtracting links intersecting with current cycling infrastructure identifies links proposed for intervention.

@@

@@

```{r potdemhistograms, fig.cap="Distribution of Potential Cycling Demand", out.width="32%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/histogram_distance_all_vs_cycling.png",
  "data/Manchester/Plots/histogram_distance_all_vs_cycling_potential.png",
  "data/Manchester/Plots/histogram_distance_cycling_potential_vs_current.png"
))
```

```{r desirefacetcycling, fig.cap="Current and Potential Cycling Demand", out.width="80%", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/desire_facet_cycling.png")
```

It should be noted that the calculations assume a future that is constrained by physical geography; i.e. we consider cycling in the traditional sense. Recently there have been various micro-mobility solutions, including e-bikes, that allow commuters to traverse longer distances and hillier roads with less effort than traditional bicycles. While these modes would probably be associated with less geographical impedance, it is beyond the scope of this work to integrate that into the analysis. Doing so is partially restricted by the lack of data on the proliferation of these modes, which raises the point that perhaps the census data category of `Bicycle' is too vague, and should be further dis-aggregated to distinguish between traditional bicycles and other forms of micro-mobility.

@@

**ADD TABLE**

@@

The weighting profiles are therefore adjusted to favor less-stressful streets (based on information from Table **REFERENCE THE TABLE**), and roads with existing cycling infrastructure. This is also in line with the creation of LTNs, as residential streets are those where motorized traffic is most likely to be banned in the creation of LTNs.

**ADD TABLE**

\noindent A weighted distance $\boldsymbol{d_{w}}$ for each road segment is calculated as following ^[The **dodgr** r package [@padgham2019dodgr] is used to route cycling demand onto the road network. The package uses the OpenStreetMaps (OSM) road network and allows the user to assign weights to roads based on their type. The routing is done based on weighted shortest paths, with the distance along each road segment being divided by a factor to obtain the weighted distance for routing. It is more intuitive to multiply when weighting a network, but the dodgr package divides by numbers between 0 and 1, which achieves the same result. For the sake of reproducibility, we stick to the convention used in the package.]:

\begin{equation}\label{eq:weight_distance}
    d_{w} = d_{unw}  \:/\:  W
\end{equation}

\noindent where $\boldsymbol{d_{unw}}$ is the unweighted distance and $\boldsymbol{W}$ is the weight from Table \ref{table:weightprofiles}.

All weights are between 0 and 1, and the values in the \textit{Weighted} profile are chosen so as to be inversely proportional to the stress level experienced by cyclists on them. The \textit{Unweighted} weighting profile is used to compare increases in route length resulting from two different approaches:

1. **Weighted**: Relatively high impedance on Primary and Trunk roads (to minimize cycling on them).
2. **Weighted\_2**: Avoiding Primary and Trunk Roads completely.

Comparing the cycling demand routed on the weighted and unweighted road network allows us to get a better understanding of the importance of different road types. In the case of Manchester, trunk roads bisect the city and are a major part of unweighted shortest paths (Figure \@ref(fig:flowsfacetunweighted)). On the other hand, cycleways are not part of unweighted shortest paths, and so very little of the cycling demand is routed through them. In the weighted network, cycleways are much better utilized, and the majority of the cycling demand passes through tertiary roads, as expected.

```{r flowsfacetunweighted, fig.cap="Flow Results Based on **Unweighted** Shortest Paths (Manchester)", out.width="90%", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_unweighted_Manchester.png")
```

```{r flowsfacetweighted, fig.cap="Flow Results Based on **Weighted** Shortest Paths (Manchester)", out.width="90%", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_weighted_Manchester.png")
```

<!-- The difference between aggregate flow on weighted and unweighted networks is dependant on the road network of the city. Comparing Manchester to Nottingham, we see that trunk roads are much more important in the former, as over 25\% of flow on the unweighted road network passes through them. For Nottingham, less than 10\% of the flow on the unweighted network passes through trunk roads, but almost 25\% of the flow passes through tertiary roads (Figure \ref{fig:perc_person-km}).  -->

The results of routing potential cycling demand on the weighted and unweighted networks are understandably quite different. From Figure \@ref(fig:flowsfacetunweighted) we can see that trunk and primary roads are the most efficient means of traversing the road network of Manchester. Both of these road types are classified as Primary A roads according to the UK Department for Transport's road classification (**Reference the table**), and are therefore part of the Primary Route Network (PRN) [@department2012guidance]. The PRN has the widest, most direct roads on the network, and carries most of the through traffic. This includes freight, with all roads in the PRN being required by law to provide unrestricted access to trucks up to 40 tonnes [@department2012guidance].

We choose to avoid routing the potential cycling demand on Primary A Roads for the following 2 reasons:

1. **Logistical Difficulty**: Changes on these roads need to be agreed upon by all affected authorities [@department2012guidance], which may prove to be difficult. 
2. **Low Traffic Neighborhoods (LTNs)**: The UK government is aiming to restrict access to motorized vehicles on residential roads to create LTNs [@departmentgearchange2020]. This is part of a policy to prevent automobile rat-running and make streets more accessible to cyclists and pedestrians. Under such a policy, Primary A roads would become even more essential for motorized traffic and it would be more difficult to reallocate road space on these roads to cyclists.

Routing potential cycling demand on a weighted network is more in line with government policy to create LTNs. Figure \@ref(fig:flowsfacetweighted) shows that routing on the weighted network significantly reduces flow on the trunk and primary roads, but does not eliminate it completely. This is intentional, as the impedance on these roads is only slightly higher than remaining road types (See Table **Reference the table**). Potential cycling demand is only routed on these roads if there are no routes through other roads that offer comparable directness.

Banning cycling flow completely on trunk and primary roads may result in excessively circuitous paths, as seen in Figure \ref{fig:boxplotcircuity}. When routing using the weighting profile in Table **Reference Table**, we see that shortest paths increase by less than 5\% on average from unweighted shortest paths, with the largest increases still below 30\%. When routing on primary and trunk roads is banned, the average increase relative to unweighted shortest paths rises to 10\%, with certain locations experiencing more significant negative effects on accessibility.

```{r boxplotcircuity, fig.cap="Effect of Banning Cyclists from Trunk and Primary Roads for all OD Pairs (Manchester)", out.width="35%", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/boxplot_weighted_unweighted_distances.png")
```

Given that cyclists will only deviate from shortest paths by a certain amount to access better cycling infrastructure (as explained in Section \@ref(what-affects-the-decision-to-cycle), allowing flow on some stretches of trunk and primary roads is necessary to insure cycling uptake and equitable access to cycling infrastructure. In its new vision for walking and cycling, the Department for Transport acknowledges that minimal segregated stretches of bicycle lanes on main roads will be necessary to avoid circuitous cycling networks [@departmentgearchange2020].

Weighting the road network also allows us to better utilize existing cycling infrastructure, as can be seen by the higher flow on cycleways in Figure \@ref(fig:flowsfacetweighted). Again, the small differences in impedance between cycleways and other road types mean that cycleways that require significant deviation are not routed on.

It should be reiterated that the weighting profile used for routing has been developed for the purposes of this study. It creates a hierarchy of road preference that is grounded in cyclist preference and government plans to create LTNs. Sensitivity analysis should be done to determine an optimal weighting profile, but given the variation in city road networks , <!-- (Figure \ref{fig:perc_person-km}) -->, these would probably require calibration to the specific city. More accurate routing could be carried out given the availability of road-level data. In such cases we would add additional impedance to specific roads, giving more useful routing results than the current methodology which considers all roads of the same type to be equivalent.

One use-case of such granular data would be to identify roads that serve schools. The Department of Transport notes that the number of school children being driven to school has trebled over the past 40 years [@departmentgearchange2020], and so having cycling infrastructure serving schools is key to achieving the government target of getting more children to cycle. This would not be difficult, as over 75\% of children in the UK live within a 15 minute cycle from their school [@departmentcycleinfradesign2020]. @goodman2019scenarios show that if dutch levels of cycling were achieved in the UK, the \% of children cycling to school could increase from 1.8\% to 41\%.

In their typology of cyclists, @dill2013four found that a majority of people who say they would never cycle had never cycled to school, whereas confident cyclists were those most likely to have cycled to school. Getting people to cycle from a young age is therefore key to achieving societal change in commuting habits.
